---
title: An analysis of weighted on-base averages with ebnm
author: Jason Willwersheid, Peter Carbonetto and Matthew Stephens
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
bibliography: ebnm.bib
vignette: >
  %\VignetteIndexEntry{An analysis of weighted on-base averages with ebnm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we illustrate the key features of ebnm in an
analysis of baseball statistics.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## The "wOBA" data set

We begin by loading and inspecting the wOBA data set, which
consists of wOBAs ("weighted on-base averages") and standard errors
for the 2022 MLB regular season:

```{r load-data}
library(ebnm)
data(wOBA)
nrow(wOBA)
head(wOBA)
```

Column "x" contains the observed wOBAs, which we interpret as
estimates of the players' "true" wOBA skill. Column "s" gives the
standard errors. (See below for background on the wOBA statistic and
details on how the standard errors were calculated.)

Next, we visualize the overall distribution of wOBAs:

```{r woba-histogram, fig.height=2, fig.width=4}
library(ggplot2)
ggplot(wOBA, aes(x = x)) +
  geom_histogram(bins = 64, color = "white",fill = "black") +
  theme_classic()
```

As the histogram shows, most players finished the season with a wOBA
between .200 and .400. A few had very high wOBAs (>.500), while others
had wOBAs at or near zero. A casual inspection of the data suggests
that players with these very high (or very low) wOBAs were simply
lucky (or unlucky). For example, the four players with the highest
wOBAs each had 16 PAs or fewer. It is unlikely that they would have
sustained this high level of production over a full season's worth of
PAs!

In contrast, Aaron Judge's production --- which included a
record-breaking number of home runs --- appears to be "real," since it
was sustained over nearly 700 PAs.  Other cases are more ambiguous:
how, for example, are we to assess Matt Carpenter, who had several
exceptional seasons between 2013 and 2018 but whose output steeply
declined in 2019--2021 before his surprising "comeback" in 2022? An
empirical Bayes analysis can help to answer this and other questions.

## The "ebnm" function

Function `ebnm()` is the main interface for fitting the empirical
Bayes normal means model; it is a "Swiss army knife" that allows for
various choices of prior family $\mathcal{G}$ as well as providing
multiple options for fitting and tuning models. For example, we can
fit a normal means model with the prior family $\mathcal{G}$ taken to
be the family of normal distributions:

```{r}
x <- wOBA$x
s <- wOBA$s
names(x) <- wOBA$Name
names(s) <- wOBA$Name
fit_normal <- ebnm(x, s, prior_family = "normal", mode = "estimate")
```

(The default behavior is to fix the prior mode at zero.
Since we certainly do not expect the distribution of true wOBA
skill to have a mode at zero, we set `mode = "estimate"`.)

We note in passing that the `ebnm` package has a second model-fitting interface, in which
each prior family gets its own function:

```{r}
fit_normal <- ebnm_normal(x, s, mode = "estimate")
```

Textual and graphical overviews of results can be obtained using, respectively, methods `summary()` and `plot()`. The summary method appears as follows:

```{r}
summary(fit_normal)
```

The `plot()` method visualizes results, comparing
the "observed" values $x_i$ (the initial wOBA estimates) against the
empirical Bayes posterior mean estimates $\hat{\theta}_i$:

```{r, fig.height=3, fig.width=3.5}
plot(fit_normal)
```

The dashed line shows the diagonal $x = y$, which makes shrinkage effects clearly visible.  In particular, the most extreme
wOBAs on either end of the spectrum are strongly shrunk towards the league
average (around .300).

Since `plot()` returns a "ggplot" object [@ggplot2], the
plot can conveniently be customized using `ggplot2` syntax. For example,
one can vary the color of the points by the number of plate
appearances:

```{r, fig.height=3, fig.width=4}
plot(fit_normal) +
  geom_point(aes(color = sqrt(wOBA$PA))) +
  labs(x = "wOBA", y = "EB estimate of true wOBA skill", 
       color = expression(sqrt(PA))) +
  scale_color_gradient(low = "blue", high = "red")
```

By varying the color of points, we see that the wOBA estimates with
higher standard errors or fewer plate appearances (blue points) tend
to be shrunk toward the league average much more strongly than wOBAs
from hitters with many plate appearances (red points).

Above, we used `head()` to view data for the first 6 hitters in the
dataset. Let's now see what the EBNM analysis suggests might be their
"true" wOBA skill. To examine the results more closely, we use the
`fitted()` method, which returns a posterior summary for each
hitter:

```{r}
print(head(fitted(fit_normal)), digits = 3)
```

The wOBA estimates of the first four ballplayers are shrunk strongly
toward the league average, reflecting the fact that these players had
very few plate appearances (and indeed, we were not swayed by their
very high initial wOBA estimates).

Carpenter had many more plate appearances (154) than these other
four players, but according to this model we should remain skeptical about
his strong performance; after factoring in the prior, we judge his
"true" performance to be much closer to the league average,
downgrading an initial estimate of .472 to the final posterior mean
estimate of .339.

## Background on the "weighted on-base average"

A longstanding tradition in empirical Bayes research is to include an
analysis of batting averages using data from Major League Baseball;
see, for example, @BrownBaseball; @JiangZhangBaseball;
@GuKoenkerBaseball. Until recently, batting averages were the most
important measurement of a hitter's performance, with the prestigious
yearly "batting title" going to the hitter with the highest
average. However, with the rise of baseball analytics, metrics that
better correlate to teams' overall run production have become
increasingly preferred. One such metric is wOBA ("weighted on-base
average"), which is both an excellent measure of a hitter's offensive
production and, unlike competing metrics such as MLB's xwOBA [@xwoba]
or Baseball Prospectus's DRC+ [@drcplus], can be calculated using
publicly available data and methods.

Initially proposed by @Tango, wOBA assigns values ("weights") to
hitting outcomes according to how much the outcome contributes on
average to run production. For example, while batting average treats
singles identically to home runs, wOBA gives a hitter more than twice
as much credit for a home run. (Note that weights are updated from year
to year, but wOBA weights for singles have remained near 0.9 for the
last several decades, while weights for home runs have hovered
around 2.0; see @fgguts.)

Given a vector of wOBA weights $\mathbf{w}$, hitter $i$'s wOBA is the
weighted average
$$
x_i = \mathbf{w}^\top \mathbf{z}^{(i)} / n_i,
$$
where $\mathbf{z}^{(i)} = (z_1^{(i)}, \ldots, z_7^{(i)})$ tallies
  outcomes (singles, doubles, triples, home runs, walks,
  hit-by-pitches and outs) over the hitter's $n_i$ plate appearances
  (PAs). Modeling hitting outcomes as i.i.d.
$$
\mathbf{z}^{(i)} \sim \text{Multinomial}(n_i, \mathbf{\pi}^{(i)}),
$$
where $\mathbf{\pi}^{(i)} = (\pi_1, \ldots, \pi_7^{(i)})$ is the vector of
"true" outcome probabilities for hitter $i$, we can regard $x_i$ as
a point estimate for the hitter's "true wOBA skill",
$$
\theta_i = \mathbf{w}^\top \mathbf{\pi}^{(i)}.
$$
Standard errors for the $x_i$'s can be estimated as
$$
s_i^2 = \mathbf{w}^\top \hat{\mathbf{\Sigma}}^{(i)} \mathbf{w}/n_i,
$$
where $\hat{\mathbf{\Sigma}}^{(i)}$ is the estimate of the covariance
matrix for the multinomial model obtained by setting $\mathbf{\pi} =
\hat{\mathbf{\pi}}$, where
$$
\hat{\mathbf{\pi}}^{(i)} = \mathbf{z}^{(i)}/n_i.
$$
(To deal with small sample sizes, we conservatively lower bound each
standard error by the standard error that would be obtained by
plugging in league-average event probabilities
$\hat{\mathbf{\pi}}_{\mathrm{lg}} = \sum_{i=1}^N \mathbf{z}^{(i)}/
\sum_{i=1}^N n_i$, where $N$ is the number of hitters in the data set.)

The relative complexity of wOBA makes it well suited for analysis via
ebnm. With batting average, a common approach is to obtain empirical
Bayes estimates using a beta-binomial model (see, for example,
@robinson). With wOBA, one can estimate hitting outcome probabilities
by way of a Dirichlet-multinomial model; alternatively, one can
approximate the likelihood as normal and fit an EBNM model directly to
the observed wOBAs. We take the latter approach.

## Bibliography
